### 2024-07-30 machine-learning.qmd
20240730 - merging, data prep, imputation

```{r}
library("LongituRF")
```

@@

```{r}
# Use common name for the same (gsis_id) ID variable
nfl_actualStats_offense_weekly <- nfl_actualStats_offense_weekly %>% 
  rename(gsis_id = player_id)

@@

@@

idSeasonWeekToMerge <- list(
  nfl_rosters_weekly %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week)),
  nfl_depthCharts %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week)),
  #nfl_actualStats_offense_weekly,
  nfl_expectedFantasyPoints_weekly %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week))
  #nfl_advancedStatsPFR_weekly,
)

# Merge data
idMerged <- idListToMerge %>% 
  reduce(full_join, by = c("gsis_id"))

idSeasonMerged <- idSeasonListToMerge %>% 
  reduce(full_join, by = c("gsis_id","season"))

idSeasonWeekMerged <- idSeasonWeekToMerge %>% 
  reduce(full_join, by = c("gsis_id","season","week"))

seasonalData <- full_join(
  idMerged,
  idSeasonMerged,
  by = "gsis_id"
)

seasonalAndWeeklyData <- full_join( # need to troubleshoot error
  seasonalData,
  idSeasonWeekMerged,
  by = c("gsis_id","season")
)
```

@@

>>>>

```{r}
library("LongituRF")
library("missForest")
library("powerjoin")
library("tidyverse")
```

@@

```{r}
# Prepare data for merging
## Use common name for the same (gsis_id) ID variable
nfl_actualStats_offense_weekly <- nfl_actualStats_offense_weekly %>% 
  rename(gsis_id = player_id)

@@

## Ensure variables with the same name have the same type
nfl_players <- nfl_players %>% 
  mutate(
    birth_date = as.Date(birth_date),
    jersey_number = as.character(jersey_number),
    gsis_it_id = as.character(gsis_it_id),
    years_of_experience = as.integer(years_of_experience))

player_stats_seasonal_offense <- player_stats_seasonal_offense %>% 
  mutate(
    birth_date = as.Date(birth_date),
    jersey_number = as.character(jersey_number),
    gsis_it_id = as.character(gsis_it_id))

nfl_rosters <- nfl_rosters %>% 
  mutate(
    draft_number = as.integer(draft_number))

nfl_rosters_weekly <- nfl_rosters %>% 
  mutate(
    draft_number = as.integer(draft_number))

nfl_depthCharts <- nfl_depthCharts %>% 
  mutate(
    season = as.integer(season))

nfl_expectedFantasyPoints_weekly <- nfl_expectedFantasyPoints_weekly %>% 
  mutate(
    receptions = as.integer(receptions))

## Rename variables
#-todo: rename variables in expected fantasy points so they don't get coalesced with actual points
nfl_draftPicks <- nfl_draftPicks %>%
  rename(
    games_career = games,
    pass_completions_career = pass_completions,
    pass_attempts_career = pass_attempts,
    pass_yards_career = pass_yards,
    pass_tds_career = pass_tds,
    pass_ints_career = pass_ints,
    rush_atts_career = rush_atts,
    rush_yards_career = rush_yards,
    rush_tds_career = rush_tds,
    receptions_career = receptions,
    rec_yards_career = rec_yards,
    rec_tds_career = rec_tds,
    def_solo_tackles_career = def_solo_tackles,
    def_ints_career = def_ints,
    def_sacks_career = def_sacks
  )

@@

idSeasonWeekListToMerge <- list(
  nfl_rosters_weekly %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week)),
  nfl_depthCharts %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week)),
  #nfl_actualStats_offense_weekly,
  nfl_expectedFantasyPoints_weekly %>% filter(!is.na(gsis_id), !is.na(season), !is.na(week))
  #nfl_advancedStatsPFR_weekly,
)

# Merge data
idMerged <- idListToMerge %>% 
  reduce(
    powerjoin::power_full_join,
    by = c("gsis_id"),
    conflict = coalesce_xy)

idSeasonMerged <- idSeasonListToMerge %>% 
  reduce(
    powerjoin::power_full_join,
    by = c("gsis_id","season"),
    conflict = coalesce_xy)

idSeasonWeekMerged <- idSeasonWeekListToMerge %>% 
  reduce(
    powerjoin::power_full_join,
    by = c("gsis_id","season","week"),
    conflict = coalesce_xy)

seasonalData <- powerjoin::power_full_join(
  idSeasonMerged,
  idMerged,
  by = "gsis_id",
  conflict = coalesce_xy
)

seasonalAndWeeklyData <- powerjoin::power_full_join(
  idSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy
)
```

### Subset to Predictor Variables and Outcome Variable {#sec-subsetToPredictorsAndOutcome}

```{r}
seasonalData %>% select_if(~class(.) == "Date")
seasonalData %>% select_if(is.character)
seasonalData %>% select_if(is.factor)
seasonalData %>% select_if(is.logical)

seasonalData_subset <- seasonalData %>% 
  dplyr::select(-c(
    birth_date, loaded, full_name, player_name, player_display_name, display_name, suffix, headshot_url, player, pos,
    gsis_id, espn_id, sportradar_id, yahoo_id, rotowire_id, pff_id, fantasy_data_id, sleeper_id, pfr_id,
    pfr_player_id, cfb_player_id, pfr_player_name, esb_id, gsis_it_id, smart_id,
    college, college_name, team_abbr, current_team_id, college_conference, draft_club, status_description_abbr,
    status_short_description, short_name, headshot, uniform_number, jersey_number, first_name, last_name,
    football_name, team
    ))
```

### Convert Character Predictors to Factors {#sec-convertCharacterToFactor}

```{r}
seasonalData_subset <- seasonalData_subset %>% mutate(across(where(is.character), as.factor))
seasonalData_subset <- seasonalData_subset %>% mutate(across(where(is.logical), as.factor))

seasonalData_subset %>% select_if(~class(.) == "Date") %>% str()
seasonalData_subset %>% select_if(is.character) %>% str()
seasonalData_subset %>% select_if(is.factor) %>% str()
seasonalData_subset %>% select_if(is.logical) %>% str()
```

### Impute the Missing Data {#sec-missingDataImputation}

```{r}
seasonalData_subset_imp <- missForest::missForest(seasonalData_subset %>% data.frame)
```

@@

```{r}
smerf <- LongituRF::MERF(
  X = seasonalData[,c("ageCentered20")] %>% as.matrix(),
  Y = seasonalData$fantasy_points,
  Z = seasonalData[,c("ageCentered20")] %>% as.matrix(),
  id = seasonalData$gsis_id,
  time = seasonalData$ageCentered20,
  mtry = 2,
  ntree = 500,
  sto = "BM")

smerf$forest # is the fitted random forest (obtained at the last iteration)
smerf$random_effects # are the predicted random effects for each individual
smerf$omega # are the predicted stochastic processes
plot(smerf$Vraisemblance) # evolution of the log-likelihood
smerf$OOB # OOB error at each iteration
```
